Cognate beginnings to bilingual lexical acquisition

### 2023-10-03 manuscript/manuscript.qmd
Ignore .tiff, .eps, and .svg

---
abstract: |
    Bilingual infants' developmental trajectories of lexical acquisition are equivalent to their monolingual peers'. This is remarkable, given the complexity of their linguistic input. Recent studies suggest that bilingual vocabulary growth is boosted by the number of cognates (form-similar translation equivalents) shared by the pair of languages being learned, and that this cognateness facilitation effect is driven by a stronger parallel activation of cognates during linguistic exposure, compared to non-cognates. The mechanisms behind this facilitation are still unclear. In this study, we propose an account of bilingual lexical acquisition in which parallel activation increases the rate at which children accumulate learning instances for words in both languages, even in fully monolingual situations. We predicted a stronger cognate facilitation for words to which children were exposed less frequently (low-exposure words), as they are co-activated by their translation more often than high-exposure words. We developed an extensive online vocabulary checklist, the Barcelona Vocabulary questionnaire (BVQ), to collect vocabulary data from 366 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We used Bayesian explanatory item response theory to model the acquisition trajectories of 604 Catalan and Spanish words. We found an interaction between exposure and cognateness, suggesting that cognateness facilitates the aquisition of low-exposure words, but not of mean exposure or high-exposure words. Overall, our findings suggest that cognateness plays a key role in bilingual lexical acquisition, and provide evidence for a frequency-mediated facilitation effect driven by parallel activation.

thanks: |
    The authors declare no conflicts of interest with regard to the funding source of this study. This study was supported by the Spanish Ministry for Science and Innovation and State Research Agency (Project PID2021- 123416NB-I00 financed by MCIN/ AEI/ 10.13039/501100011033 / FEDER, UE) and the Economic and Social Research Council (ESRC) (ES/S010947/1, UK). GGC was supported by a FPI research contract (PRE2019-088165). DAV was supported by the European Union's Horizon 2023 research and innovation program under Marie Skłodowska–Curie Grant (765556) and a postdoctoral fellowship from the Foundation for Science and Technology of Portugal (UIDB/00214/2020). IC was supported by the Investigo program funded by the European Union's NextGenerationEU (NGEU) recovery plan. NSG was supported by an ICREA Academia award from the Catalan Institution for Research and Advanced Studies (ICREA). We are grateful to Chiara Santolin, Ege E. Özer, and the rest of the Speech Acquisition and Perception research group, and to Alicia Franco-Martínez and Cristina Rodríguez-Prada, for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.
---

@@

@@

@floccia2018introduction published an impressive study comparing vocabularies of children learning several language pairs differing in their percentage of cognates. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi or Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The authors calculated the average lexical overlap between the words in each of these additional languages and their translation equivalents in English. Lexical overlap was calculated in terms of phonological similarity (described below) and it was taken as a proxy of the degree of cognateness between each pair of languages. Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (languages with a high phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (languages with a low phonological overlap) were able to produce in Mandarin. @blom2020crosslanguage, @bosma2019longitudinal, and @gampe2021does reported similar results, providing converging evidence of a facilitatory effect of a lower language distance (i.e., higher degre of cognateness) on vocabulary size.

A second set of studies suggested that cognates are overrepresented in bilinguals’ early lexicon. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months and found that cognates represented a larger proportion of vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual child produced cognates earlier than non-cognates, on average. But the high proportion of cognates in the vocabulary of the participants in these two studies may not necessarily evidence of a facilitation effect of cognateness, but rather of simply the high proportion of cognates present in the pair of languages being learned. For instance, if two given languages share a high proportion of cognates like 70%, the vocabulary contents of children learning both languages should, in principle, approximate such proportion of cognates, even in the absence of a cognateness facilitation effect. More recently, @mitchell2022cognates addressed this issue in a longitudinal study. The authors collected expressive vocabulary data of 47 16- to 30-month-old French-English bilinguals living in Canada, in both languages. They created two lists of translation equivalents: one made of 131 cognates, and one made of 406 non-cognates. The proportion of words that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list across ages. Critically, this difference persisted after both lists were matched in size, controlling their semantic category (i.e., furniture, animals, food were similarity represented in both lists) and age-of-acquisition norms (an index of word difficulty). Taken together, the results of these two lines of research have been considered as supporting to the hypothesis that phonological similarity (as reflected in cognateness) plays a facilitation role in bilingual word acquisition.

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual contexts. Evidence with adult bilinguals supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @hoshino2008cognate; @morford2011when; @spivey1999cross; @shook2012bimodal; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*). Obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. More recntly, evidence of parallel activation has been reported in bilingual toddlers and children too [@vonholzen2012language; @floccia2020translation; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @poulin-dubois2013lexical]. Although there is a consensus on the role of parallel activation in bilinguals' lexical processing and acquisition, previous studies do not address its influence on the learning trajectories of words. Results are aggregated across words and provide no information about the specific dynamics of how parallel activation influences word learning. This is the goal of the present research.

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, to the extent that such translation equivalent is co-activated. We use the term *learning instance* in the fashion of accumulator models of language acquisition; as an exposure to a word-form that constitutes an opportunity for the child to accumulate information about the word. We do not assume if a learning instance is a discrete or a continuous unit of accumulation of information. We consider that a learning instance of a word is an exposure to its (phonological) form if the resulting strength of activation of its representation in the child's lexicon reaches some thoretical threshold that leads to word-form recognition. This activation may result from the infant being exposed to the actual word-form, or the result of activation spreading through phonological or semantic links across lexical representation, as in the case of parallel activation. The strength of this co-activation is proportional to the phonological similarity between the two translation equivalents; given that cognates share higher phonological similarity than non-cognates, the former should be co-activated more strongly than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across languages: words from the lower-exposure language would receive stronger activation from words in the higher-exposure language than vice versa. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than words from the higher-exposure language. This asymmetric cross-language activation would be consistent with previous reports of larger priming effects from the dominant to the non-dominant language.

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}/--/\textipa{"ga.to}/ [*cat*], which are phonologically very similar. When the child listens to /\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in parallel. If the child has already formed a form-meaning association for both word-forms, parallel activation may result from the activation of their common concept or from activation spreading throught phonological similarity. We assume semantic co-activation to be constant across cognate and non-cognate translation equivalents, and focus on phonological co-activation as an additional source of activation that affects cognates more strongly than non-cognates. Therefore, this exposure will count as a learning instance for both co-activated forms. The case of the non-cognate translation equivalent /\textipa{"gos-"pe.ro}/ [*dog*] would be different. Given the low phonological similarity between both word-forms, an exposure to \textipa{"gos} will result in a weak activation of /\textipa{"pe.ro}/ leading to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for /\textipa{"pe.ro}/. While /\textipa{"gat}/--/\textipa{"ga.to}/ will benefit from phonological co-activation, /\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives linguistic input from one of the languages more often than from the other, this effect might affect each form of the cognate translation equivalent differently. For instance, if the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will activate /\textipa{"ga.to}/ more often than vice versa. Ultimately, /\textipa{"ga.to}/ will benefit more strongly from its cognate status than /\textipa{"gat}/, as it receives additional learning instances from its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children. We adopted a Bayesian explanatory item response theory approach to model the probability of acquisition of `r nrow(distinct(items, language, item))` Catalan and Spanish nouns included in the vocabulary checklist. Words were considered as acquired if caregivers reported such word to be understood (comprehension) or understood and said (production) by their child. We estimated the impact of several predictors of interest on the probability of acquisition, including participants’ age and rate of exposure to the word-form, and the cognate status of the word-form. As described in the methods section, rate of exposure was a composite measure taking into account participant’ language exposure and word’s lexical frequency. We predicted an interaction between cognate status and word-form exposure rate  in which the probability of comprehension is higher for low-exposure cognate words, but not for high-exposure cognate words.

@@

All materials, data, and reproducible code can be found at the OSF ([https://osf.io/hy984/](https://osf.io/hy984/)) and GitHub ([https://github.com/gongcastro/cognate-beginnings](https://github.com/gongcastro/cognate-beginnings)) repositories. This study was conducted according to guidelines laid down in the Declaration of Helsinki, and was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I.

@@

n_items_version <- bvq_data$pool |>
    unnest(version) |>
    count(version, language) |>
    pivot_wider(
        names_from = language,
        values_from = n,
        names_repair = janitor::make_clean_names
    ) |>
    mutate(version = tolower(version))

```

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire (BVQ) [@garcia-castro2023bvq]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the `formr` platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted in two lists of words: one in Catalan and one in Spanish. Both lists included items from a diverse sample of `r n_categories` semantic or functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for *dog*), roughly following a one-to-one mapping. Some of the words in Catalan did not have a clear translation or had more than one possible translation in Spanish, and vice versa, therefore the unequal number of words included in the two lists (see @tbl-items for a summary of the questionnaire items).

@@

sum_rows <- function(x) {
    if (is.numeric(x)) {
        sum(x)
    } else if (is.list(x)) {
        ""
    } else {
        "N"
    }
}

sums_row <- tbl_data |>
    summarise(
        semantic_category = "N",
        across(a:d, sum),
        items = "-"
    )

@@

kable(tbl_data,
    format = "latex",
    digits = 1,
    booktabs = TRUE,
    col.names = col_names,
    align = "lccccl"
) |>
    row_spec(nrow(tbl_data) - 1,
        hline_after = TRUE
    )
```

For each word included in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Given the large number of words in the vocabulary checklists, we created four different subsets of the complete list of items: A, B, C, and D. Each subset contained a random but representative sub-sample of the items from the complete list (see @tbl-items). Semantic or functional categories with less than 16 items---thus resulting in less than four items after dividing it in four lists---were not divided: all of their items were included in the four lists. Items that were part of the trial lists of some ongoing experiments in the lab were also included in all versions. The resulting reduced list contained between `r min(n_items_version$catalan)` and  `r max(n_items_version$catalan)` Catalan words, and between `r min(n_items_version$spanish)` and  `r max(n_items_version$spanish)` Spanish words. Participants were randomly allocated into one of the four subsets.

@@

participants_edu <- table(participants$edu_parent)
```

We collected `r n_responses` responses to the questionnaire from `r sum(n_participants)` distinct children from the Metropolitan Area of Barcelona between Marth 30th, 2020 and October 31th, 2022: `r n_participants[1]` of those participants participated once, `r n_participants[2]` twice,  `r n_participants[3]` three times, and  `r n_participants[4]` four times. Recurrent participants provided responses with a minimum of `r as.numeric(diff_responses$diff_min)` days between responses, and a maximum of `r as.numeric(diff_responses$diff_max)` days, and were always allocated to the same questionnaire list (A, B, C, or D). Participants were part of the database of the Laboratori de Recerca en Infància (Universitat Pompeu Fabra) and were contacted by e-mail or phone if their child was aged between `r floor(min(responses$age))` and `r ceiling(max(responses$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan (see @tbl-participants for a more detailed description of the sample). In total, `r n_trilingual$n[2]` participants (`r scales::percent(n_trilingual$prop[2], accuracy = 0.01)`) participants were reported to be exposed in less than 10% to a third language other than Catalan and Spanish. All families provided informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a token of appreciation.

```{r tbl-participants}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan."
#| tbl-note: "For participants exposed to Catalan and Spanish exclusively, the proportion of Catalan exposure shown in the table is complementary to the degree of exposue to Spanish. For participants exposed to a third languages this proportion is not complementary unless one adds the degree of exposed to the third language, which never exceeded 10%. Participants with 100\\%, 0\\%, and 50\\% Catalan exposure would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively."
ftn <- ""

@@

kable(tbl_data,
    format = "latex",
    digits = 1,
    booktabs = TRUE,
    col.names = col_names,
    align = "lcccccc"
) |>
    add_header_above(c(
        " " = 1,
        "Age (months)" = ncol(tbl_data) - 1
    )) |>
    row_spec(nrow(tbl_data) - 1,
        hline_after = TRUE
    )
```

>>>>

---
abstract: |
    Bilingual infants' developmental trajectories of lexical acquisition are equivalent to their monolingual peers'. This is remarkable, given the complexity of their linguistic input. Recent studies suggest that bilingual vocabulary growth is boosted by the number of cognates (form-similar translation equivalents) shared by the pair of languages being learned, and that this cognateness facilitation effect is driven by a stronger parallel activation of cognates during linguistic exposure, compared to non-cognates. The mechanisms behind this facilitation are still unclear. In this study, we propose an account of bilingual lexical acquisition in which parallel activation increases the rate at which children accumulate learning instances for words in both languages, even in fully monolingual situations. We predicted a stronger cognate facilitation for words to which children were exposed less frequently (low-exposure words), as they are co-activated by their translation more often than high-exposure words. We developed an extensive online vocabulary checklist, the Barcelona Vocabulary questionnaire (BVQ), to collect vocabulary data from 369 Catalan-Spanish bilingual toddlers aged 12 to 32 months. We used Bayesian explanatory item response theory to model the acquisition trajectories of 604 Catalan and Spanish words. We found an interaction between exposure and cognateness, suggesting that cognateness facilitates the aquisition of low-exposure words, but not of mean exposure or high-exposure words. Overall, our findings suggest that cognateness plays a key role in bilingual lexical acquisition, and provide evidence for a frequency-mediated facilitation effect driven by parallel activation.

thanks: |
    The authors declare no conflicts of interest with regard to the funding source of this study. This study was supported by the Spanish Ministry for Science and Innovation and State Research Agency (Project PID2021- 123416NB-I00 financed by MCIN/ AEI/ 10.13039/501100011033 / FEDER, UE) and the Economic and Social Research Council (ESRC) (ES/S010947/1, UK). GGC was supported by a FPI research contract (PRE2019-088165). DAV was supported by the European Union’s Horizon2020 research and innovation program under Marie Skłodowska–Curie Grant (765556) and partially supported by Portuguese national funds through the Foundation for Science and Technology, under project UIDB/00214/2020, awarded to the Center of Linguistics of the University of Lisbon. IC was supported by the Investigo program funded by the European Union's NextGenerationEU (NGEU) recovery plan. NSG was supported by an ICREA Academia 2019 award from the Catalan Institution for Research and Advanced Studies (ICREA). We are grateful to Chiara Santolin, Ege E. Özer, and the rest of the Speech Acquisition and Perception research group, and to Alicia Franco-Martínez and Cristina Rodríguez-Prada, for their helpful feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado for their technical support, and Cristina Dominguez and Katia Pistrin for their efforts in recruiting infants. We also thank all families and infants who participated in the experiments.
---

@@

clrs <- c("#003f5c", "#58508d", "#bc5090", "#ff6361", "#ffa600")

@@

@floccia2018introduction published an impressive study comparing vocabularies of children learning several language pairs differing in their percentage of cognates. The authors collected vocabulary data on word comprehension and production from 372 24-month-old bilingual toddlers living in the United Kingdom who were learning English and an additional language. The additional language was one of 13 typologically diverse languages: Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi or Urdu, Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The authors calculated the average lexical overlap between the words in each of these additional languages and their translation equivalents in English. Lexical overlap was calculated in terms of phonological similarity (described below) and it was taken as a proxy of the degree of cognateness between each pair of languages. Floccia and co-workers reported an increase in vocabulary size in the additional language (i.e., not English) associated with an increase in the average phonological similarity between the translation equivalents of each language pair. For example, English-Dutch bilinguals (languages with a high phonological overlap), were able to produce more Dutch words than English-Mandarin bilinguals (languages with a low phonological overlap) were able to produce in Mandarin. @blom2020crosslanguage, @bosma2019longitudinal, and @gampe2021does reported similar results, providing converging evidence of a facilitatory effect of a lower language distance (i.e., higher degree of cognateness) on vocabulary size.

A second set of studies suggested that cognates are overrepresented in bilinguals’ early lexicon. @bosch2014first collected parental reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged 18 months and found that cognates represented a larger proportion of vocabulary than non-cognates. @schelletter2002effect provided converging evidence from a longitudinal single-case study, in which an English-German bilingual child produced cognates earlier than non-cognates, on average. But the high proportion of cognates in the vocabulary of the participants in these two studies may not necessarily evidence of a facilitation effect of cognateness, but rather of simply the high proportion of cognates present in the pair of languages being learned. For instance, if two given languages share a high proportion of cognates like 70%, the vocabulary contents of children learning both languages should, in principle, approximate such proportion of cognates, even in the absence of a cognateness facilitation effect. More recently, @mitchell2022cognates addressed this issue in a longitudinal study. The authors collected expressive vocabulary data of 47 16- to 30-month-old French-English bilinguals living in Canada, in both languages. They created two lists of translation equivalents; one made of 131 cognates, and one made of 406 non-cognates. Across ages, the proportion of words that children were reportedly able to produce was higher in the cognate lists than in the non-cognate list. Critically, this difference persisted after both lists were matched in size, controlling their semantic category (i.e., furniture, animals, food were similarity represented in both lists) and age-of-acquisition norms (an index of word difficulty). Taken together, the results of these two lines of research support the hypothesis that phonological similarity (as reflected in cognateness) plays a facilitation role in bilingual word learning.

Parallel activation of bilinguals’ lexicons has been proposed as the underlying mechanism for such facilitatory effect [e.g., @floccia2018introduction; @mitchell2022cognates]. The parallel activation hypothesis stems from the language non-selective account of lexical access, which suggests that bilinguals activate both languages simultaneously during language processing, even in fully monolingual contexts. Evidence with adult bilinguals supporting the language-non selective account of lexical access has been reported for language comprehension and production, across the auditory and visual (reading and signing) modalities [@gimeno-martinez2021crosslinguistic; @hoshino2008cognate; @morford2011when; @spivey1999cross; @shook2012bimodal; see @kroll2017bilingual for review]. One of the clearest pieces of evidence of parallel activation was provided by @costa2000cognate. In this study, Spanish monolinguals and Catalan-Spanish bilingual adults were asked to name pictures of common objects in Spanish. In half of the trials, the object labels were cognates in Spanish and Catalan (*árbol*-*arbre*, translations of *tree*), whereas in the other half of the trial labels were non-cognates (*mesa*-*taula*, translations of *table*). Obviously, such distinction was only relevant for bilinguals. Bilinguals named cognate pictures faster than non-cognate pictures, even after adjusting for the lexical frequency of the items. In contrast, Spanish monolinguals, who were unfamiliar with the Catalan translations of the Spanish words they uttered, showed equivalent naming times for the two types of stimuli. The authors interpreted the difference between cognates and non-cognates in bilinguals as reflecting the additional phonological activation that cognate words would receive from their translation equivalents (due to language non-selective activation of bilinguals’ lexicons). These results showed that bilinguals’ Catalan phonology was activated during the production of Spanish words, facilitating the naming of cognate pictures. More recently, evidence of parallel activation has been reported in bilingual toddlers and children too [@vonholzen2012language; @floccia2020translation; @vonholzen2019impact; @bosma2020cognate; @poarch2012crosslanguage; @poulin-dubois2013lexical]. Although there is a consensus on the role of parallel activation in bilinguals' lexical processing and acquisition, previous studies do not address its influence on the learning trajectories of individual words. Results are aggregated across words and provide no information about the specific dynamics of how parallel activation influences word learning. This is the goal of the present research.

We propose an account in which a learning instance for a word may also represent a learning instance for its translation equivalent, to the extent that such translation equivalent is co-activated. We use the term *learning instance* in the fashion of accumulator models of language acquisition; as an exposure to a word-form that constitutes an opportunity for the child to accumulate information about the word. We do not assume if a learning instance is a discrete or a continuous unit of accumulation of information. We consider that a learning instance of a word is an exposure to its (phonological) form if the resulting strength of activation of its representation in the child's lexicon reaches some thoretical threshold that leads to word-form recognition. This activation may result from the infant being exposed to the actual word-form, or the result of activation spreading through phonological or semantic links across lexical representation, as in the case of parallel activation. The strength of this co-activation is proportional to the phonological similarity between the two translation equivalents; given that cognates share higher phonological similarity than non-cognates, the former should be co-activated more strongly than the latter. This should lead to a faster accumulation of learning instances for cognates, compared to non-cognates. Parallel activation would allow bilingual children to accumulate learning instances for words in both languages even during fully monolingual situations, but the impact of this mechanism would be asymmetric across languages: words from the lower-exposure language would receive stronger activation from words in the higher-exposure language than vice versa. Therefore, the acquisition of words from the lower-exposure language would benefit more strongly from their cognate status than the acquisition of words from the higher-exposure language. This asymmetric cross-language activation would be consistent with previous reports of larger priming effects from the dominant to the non-dominant language [e.g., @grainger1998masked].

Consider the example of the Catalan-Spanish cognate translation equivalent /\textipa{"gat}/--/\textipa{"ga.to}/ [*cat*], which are phonologically very similar. When the child listens to /\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in parallel. If the child has already formed a form-meaning association for both word-forms, parallel activation may result from the activation of their common concept or from activation spreading throught phonological similarity. We assume semantic co-activation to be constant across cognate and non-cognate translation equivalents, and focus on phonological co-activation as an additional source of activation that affects cognates more strongly than non-cognates. Therefore, this exposure will count as a learning instance for both co-activated forms. The case of the non-cognate translation equivalent /\textipa{"gos}/--/\textipa{"pe.ro}/ [*dog*] would be different. Given the low phonological similarity between both word-forms, an exposure to \textipa{"gos} will result in a weak activation of /\textipa{"pe.ro}/ leading to such exposure counting as a learning instance for /\textipa{"gos}/ (which the child was exposed to), but not for /\textipa{"pe.ro}/. While /\textipa{"gat}/--/\textipa{"ga.to}/ will benefit from phonological co-activation, /\textipa{"gos}--\textipa{"pe.ro}/ will not. If the child receives linguistic input from one of the languages more often than from the other, this effect might affect each form of the cognate translation equivalent differently. For instance, if the child receives a larger amount of Catalan input than Spanish input, they will encounter the Catalan form /\textipa{"gat}/ more frequently than the Spanish form /\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will activate /\textipa{"ga.to}/ more often than vice versa. Ultimately, /\textipa{"ga.to}/ will benefit more strongly from its cognate status than /\textipa{"gat}/, as it receives additional learning instances from its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production and comprehension from a large sample of bilingual Catalan-Spanish children. We adopted a Bayesian explanatory item response theory approach to model the probability of acquisition of `r nrow(distinct(items, language, item))` Catalan and Spanish nouns included in the vocabulary checklist. Words were considered as acquired if caregivers reported such word to be understood (comprehension) or understood and said (production) by their child. We estimated the impact of several predictors of interest on the probability of acquisition, including participants’ age and rate of exposure to the word-form, and the cognate status of the word-form. As described in the Methods section, rate of exposure was a composite measure taking into account participant’ language exposure and word’s lexical frequency. We predicted an interaction between cognate status and word-form exposure rate in which the probability of comprehension is higher for low-exposure cognate words, but not for high-exposure cognate words.

@@

All materials, data, and reproducible code can be found at the OSF ([https://osf.io/hy984/](https://osf.io/hy984/)) and GitHub ([https://github.com/gongcastro/cognate-beginnings](https://github.com/gongcastro/cognate-beginnings)) repositories. For reproducibility, a Docker image of the RStudio session is available on DockerHub([https://hub.docker.com/repository/docker/gongcastro/cognate-beginnings/](https://hub.docker.com/repository/docker/gongcastro/cognate-beginnings/)). This study was conducted according to guidelines laid down in the Declaration of Helsinki, and was approved by the Drug Research Ethical Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I.

@@

n_items_version <- bvq_data$pool |>
    unnest(version) |>
    count(version, language) |>
    pivot_wider(
        names_from = language,
        values_from = n,
        names_repair = janitor::make_clean_names
    ) |>
    mutate(version = tolower(version))
```

To collect vocabulary data from participants, we created an *ad hoc* questionnaire: the Barcelona Vocabulary Questionnaire (BVQ) [@garcia-castro2023bvq]. This questionnaire was inspired by the MacArthur-Bates Communicative Development Inventory [@fenson2007macarthurbates] and its adaptations to other languages, and was implemented on-line using the `formr` platform [@arslan2020formr]. This questionnaire is structured in three blocks: (1) a language exposure questionnaire, (2) a demographic survey, and (3) two vocabulary checklists. Vocabulary checklists followed a similar structure as the Oxford Communicative Developmental Inventory [@hamilton2000infant] and consisted of two lists of words, one in Catalan and one in Spanish. Both lists included items from a diverse sample of `r n_categories` semantic or functional categories. The Catalan checklist contained `r n_item_language["catalan"]` items and the Spanish checklist contained `r n_item_language["spanish"]`. Items in one language were translation equivalents of the items in the other language (e.g., the same participant responded to both *gos* and *perro*, Catalan and Spanish for *dog*), roughly following a one-to-one mapping. Some of the words in Catalan did not have a clear translation or had more than one possible translation in Spanish, and vice versa, therefore the unequal number of words included in the two lists.

@@

sum_rows <- function(x) {
    if (is.numeric(x)) {
        sum(x)
    } else if (is.list(x)) {
        ""
    } else {
        "*N*"
    }
}

sums_row <- tbl_data |>
    summarise(
        semantic_category = "*N*",
        across(a:d, sum),
        items = "-"
    )

@@

kable(tbl_data,
    format = "latex",
    digits = 1,
    booktabs = TRUE,
    escape = FALSE,
    col.names = col_names,
    align = "lccccl"
) |>
    row_spec(nrow(tbl_data) - 1,
        hline_after = TRUE
    )
```

For each word included in the vocabulary checklists, we asked parents to report whether their child was able to understand it, understand *and* say it, or did not understand or say it (checked out by default). Given the large number of words in the vocabulary checklists, we created four different subsets of the complete list of items (A, B, C, and D) Each subset contained a random but representative sub-sample of the items from the complete list (see @tbl-items). Semantic or functional categories with less than 16 items---thus resulting in less than four items after dividing it in four subsets---were not divided: all of their items were included in the four subsets. Items that were part of the trial lists of some ongoing experiments in the lab were also included in all versions. The resulting reduced list contained between `r min(n_items_version$catalan)` and  `r max(n_items_version$catalan)` Catalan words, and between `r min(n_items_version$spanish)` and  `r max(n_items_version$spanish)` Spanish words.

@@

participants_edu <- table(participants$edu_parent)

participants_age <- summarise(participants,
                              across(age, lst(mean, median, sd, min, max)))

participants_sex <- c(table(distinct(participants, id, sex)$sex),
                      missing = sum(is.na(distinct(participants, id, sex)$sex)))
```

We collected `r n_responses` questionnaire responses from `r sum(n_participants)` distinct children (`r participants_sex[["female"]]` female, `r participants_sex[["male"]]` male, `r participants_sex[["missing"]]` not reported, mainly White). Participants were aged 12-32 months (*M* = `r round(participants_age$age_mean, 2)`, *SD* = `r round(participants_age$age_sd, 2)`, *Range* = `r round(participants_age$age_min, 2)`-`r round(participants_age$age_max, 2)`). Of those participants, `r english::words(n_participants[4])` participated four times, `r english::words(n_participants[3])` participated three times, `r n_participants[2]` participated twice, and `r n_participants[1]` participated once. Recurrent participants provided responses with a minimum of `r as.numeric(diff_responses$diff_min)` days between responses, and a maximum of `r as.numeric(diff_responses$diff_max)` days. Participants were randomly allocated into one of the four questionnaire subsets (A, B, C, or D). Each participant was always allocated to the same subset.

Participants were residents in the Metropolitan Area of Barcelona (Catalonia, Spain). Data collection took place between March 30th, 2020 and October 31th, 2022. Participants were part of the database of the Laboratori de Recerca en Infància at the Universitat Pompeu Fabra, and were contacted by e-mail or phone if their child was aged between `r floor(min(responses$age))` and `r ceiling(max(responses$age))` months, and had not been reported to be exposed more than 10% of the time to a language other than Spanish or Catalan (see @tbl-participants for a more detailed description of the sample). In total, `r n_trilingual$n[2]` participants (`r scales::percent(n_trilingual$prop[2], accuracy = 0.01)`) participants were reported to be exposed in less than 10% to a third language other than Catalan and Spanish. All families provided informed consent before participating. Upon consent, families were sent a link to the questionnaire via e-mail, which they filled from a computer, laptop, or mobile device. Filling the questionnaire took 30 minutes approximately. After completion, families were rewarded with a token of appreciation.

```{r tbl-participants}
#| label: tbl-participants
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: "Participant sample size by age and degree of exposure to Catalan."
ftn <- ""

@@

kable(tbl_data,
    format = "latex",
    digits = 1,
    booktabs = TRUE,
    col.names = col_names,
    align = "lcccccc"
) |>
    add_header_above(c(
        " " = 1,
        "Age (months)" = ncol(tbl_data) - 1
    )) |>
    row_spec(nrow(tbl_data) - 1,
        hline_after = TRUE
    ) |>
    add_footnote("For participants exposed to Catalan and Spanish exclusively, the proportion of Catalan exposure shown in the table is complementary to the degree of exposue to Spanish. For participants exposed to a third languages this proportion is not complementary unless one adds the degree of exposed to the third language, which never exceeded 10\\%. Participants with 100\\%, 0\\%, and 50\\% Catalan exposure would be traditionaly classified as Catalan monolingual, Spanish monolingual, and Catalan-Spanish bilingual, respectively.", notation = "none", escape = FALSE)
```

We used the highest self-reported educational attainment of parents or caregivers as a proxy of participants' socio-economic status (SES). This information was provided by each parent or caregiver by selecting one of six possible alternatives in line with the current educational system in Spain: *sense escolaritzar/sin escolarizar* [no education], *educació primària/educación primaria* [primary school], *educació secundària/educación secundaria* [secondary school], *batxillerat/bachillerato* [complementary studies/high school], *cicles formatius/ciclos formativos* [vocational training], and *educació universitària/educación universitaria* [university degree]. Most families reported university studies (`r participants_edu["University"]`, `r scales::percent(participants_edu["University"]/sum(participants_edu))`), followed by families where the highest educational attainment were vocational studies (`r table(participants$edu_parent)["Vocational"]`, `r scales::percent(participants_edu["Vocational"]/sum(participants_edu))`), secondary education (`r participants_edu["Secondary"]`, `r scales::percent(participants_edu["Secondary"]/sum(participants_edu))`), complementary studies (`r participants_edu["Complementary"]`,  `r scales::percent(participants_edu["Complementary"]/sum(participants_edu))`), primary education (`r participants_edu["Primary"]`, <1%), and no formal education (`r participants_edu["No education"]`, <1%).
